---
title: "How Literacy and Age at Marriage Shape Family Size in Portugal, 1979"
subtitle: "A Statistic Research Using Poisson Regression Models "
author: 
  - Shanjie Jiao
  - Xizi Sun
thanks: "Code and data are available at: https://github.com/Jie-jiao05/Family_Size."
date: today
date-format: long
abstract: "TBD. IDK. Smash Them"
format:
  pdf:
    fig-pos: H
toc: true
number-sections: true
bibliography: references.bib

---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(dplyr)
library(here)
library(readr)
library(lubridate)
library(rvest)
library(palmerpenguins)
library(maps)
library(ggplot2)
library(knitr)
library(arrow)
library(rstanarm)
library(plotly)
library(patchwork)
library(tidyr)
library(bayesplot)
library(gridExtra)
library(kableExtra)
library(broom.mixed)
library(modelsummary)

data <- read.csv("~/Family_Size/Data/cleaned_portugal.csv")
```

# Introduction
## Why this topic is important
## 3 citation relate with our research 
## Our objective, goal, how the citation related to, what we gonna do
## Brief introduction of our research


# Method
## Use GLM, possion,   (reason), what is x how to interoperate

# Result
## Data Description
This dataset originates from the 1979-80 Portuguese Fertility Survey [@PortugalWFS], conducted as part of the World Fertility Survey (WFS) program and archived by Princeton University’s Office of Population Research (OPR).

The Dataset includes 5,148 observations and 8 variables, covering factors such as literacy, marriage age, pregnancies, children, and regional demographics. @tbl-summary it shows a summary information of our dataset.

```{r}
#| label: tbl-summary
#| tbl-cap: Dataset Summary
#| echo: false
#| warning: false
#| message: false

summary_table <-summary(data)
summary_table 
```

## Model
```{r}
# Load necessary libraries
library(dplyr)
library(MASS)

# Read the cleaned dataset
data <- read.csv("~/Family_Size/Data/cleaned_portugal.csv")

# Ensure monthsSinceM is positive to prevent log(0) errors
data <- data %>%
  filter(monthsSinceM > 0) %>%
  mutate(log_monthsSinceM = log(monthsSinceM))  # Compute log offset

# Fit Poisson regression model with offset
poisson_model <- glm(children ~ ageMarried + literacy, 
                     family = poisson(link = "log"), 
                     offset = log_monthsSinceM, 
                     data = data)

# Display model summary
summary(poisson_model)

```
## Model selection (use model with offset)
## Use likelyhood rario test to do the performance ,  to see the efficiency between model
## Test overdispersion

# Conclusion

# Appendix {-}

\newpage

# References {-}
