---
title: "How Literacy and Age at Marriage Shape Family Size in Portugal, 1979"
subtitle: "A Statistic Research Using Poisson Regression Models "
author: 
  - Shanjie Jiao
  - Xizi Sun
thanks: "Code and data are available at: https://github.com/Jie-jiao05/Family_Size."
date: today
date-format: long
abstract: "TBD. IDK. Smash Them"
format:
  pdf:
    fig-pos: H
toc: true
number-sections: true
bibliography: references.bib

---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(dplyr)
library(here)
library(readr)
library(lubridate)
library(rvest)
library(palmerpenguins)
library(maps)
library(ggplot2)
library(knitr)
library(arrow)
library(rstanarm)
library(plotly)
library(patchwork)
library(tidyr)
library(bayesplot)
library(gridExtra)
library(kableExtra)
library(broom.mixed)
library(modelsummary)

data <- read.csv("~/Family_Size/Data/cleaned_portugal.csv")
```

# Introduction
## Why this topic is important
## 3 citation relate with our research 
## Our objective, goal, how the citation related to, what we gonna do
## Brief introduction of our research


# Method
## Use GLM, possion,   (reason), what is x how to interoperate

# Result
## Data Description
This dataset originates from the 1979-80 Portuguese Fertility Survey [@PortugalWFS], conducted as part of the World Fertility Survey (WFS) program and archived by Princeton University’s Office of Population Research (OPR).

The Dataset includes 5,148 observations and 8 variables, covering factors such as literacy, marriage age, pregnancies, children, and regional demographics. @tbl-summary it shows a statistical summary of our dataset.

To analyze the relationship between family size, literacy, and age at marriage, the variable Literacy was categorized as a dummy variable, assigning values 0 and 1 to illiterate and literate individuals, respectively. The analysis primarily focused on demographic factors, economic influences, and regional differences, aiming to assess their impact on family size. Additionally, an offset variable accounting for the duration of marriage was included to control for varying exposure times.

```{r}
#| label: fig-dist
#| fig-cap: Distribution of Key Variable
#| echo: false
# Load necessary libraries
library(ggplot2)
library(readr)
library(dplyr)
library(tidyr)
library(ggpubr)  # Provides ggarrange() function

# Assuming 'data' is your dataset
# Convert literacy to a factor for better visualization
data$literacy <- as.factor(data$literacy)

# Define an extra-small theme for all plots
tiny_theme <- theme_minimal(base_size = 8) +
  theme(
    plot.title = element_text(size = 8, face = "bold"),
    axis.text.x = element_text(angle = 30, hjust = 1, size = 6),
    axis.text.y = element_text(size = 6),
    axis.title = element_text(size = 7)
  )

# Histogram of Age at Marriage
p1 <- ggplot(data, aes(x = ageMarried)) +
  geom_bar(fill = "gray", color = "black", alpha = 0.7) +
  labs(title = "Distribution of Age at Marriage", x = "Age at Marriage", y = "Count") +
  tiny_theme

# Histogram of Literacy Levels
p2 <- ggplot(data, aes(x = literacy)) +
  geom_bar(fill = "gray", color = "black", alpha = 0.7) +
  labs(title = "Distribution of Literacy Levels", x = "Literacy (1 = Literate, 0 = Illiterate)", y = "Count") +
  tiny_theme

# Boxplot: Age at Marriage vs. Number of Children
p3 <- ggplot(data, aes(x = ageMarried, y = children)) +
  geom_boxplot(fill = "gray", outlier.color = "red", outlier.shape = 16) +
  labs(title = "Age at Marriage vs. Number of Children", x = "Age at Marriage", y = "Number of Children") +
  tiny_theme

# Boxplot: Literacy vs. Number of Children
p4 <- ggplot(data, aes(x = literacy, y = children)) +
  geom_boxplot(fill = "gray", outlier.color = "red", outlier.shape = 16) +
  labs(title = "Literacy vs. Number of Children", x = "Literacy (1 = Literate, 0 = Illiterate)", y = "Number of Children") +
  tiny_theme

# Ensure dataset is in data frame format
data <- as.data.frame(data)

# Select only existing region columns dynamically
region_cols <- intersect(colnames(data), c("region10_20k", "region20k_", "regionlisbon", "regionlt10k", "regionporto"))

# Ensure at least one region column exists before proceeding
if (length(region_cols) > 0) {
  region_counts <- data[, region_cols]
  region_counts <- colSums(region_counts, na.rm = TRUE)
  region_counts <- data.frame(Region = names(region_counts), Count = region_counts)

  # Barplot of Region Distribution
  p5 <- ggplot(region_counts, aes(x = Region, y = Count, fill = Region)) +
    geom_bar(stat = "identity", color = "black", alpha = 0.7) +
    labs(title = "Distribution of Regions", x = "Region", y = "Count") +
    tiny_theme +
    theme(legend.position = "bottom")
} else {
  print("No valid region columns found.")
}

# Arrange all plots in a grid with correct rendering
if (exists("p5")) {
  ggarrange(
    ggarrange(p1, p2, p3, p4, ncol = 2, nrow = 2),
    p5,
    ncol = 1,
    nrow = 2,
    heights = c(2, 1)  # Adjust the heights as needed
  )
} else {
  ggarrange(p1, p2, p3, p4, ncol = 2, nrow = 2)
}



```
@fig-dist illustrates the distribution of key variables and their relationship with family size. In general, the age of marriage is normally distributed within the range of this dataset, with a slight right skew, and the peak occurs at 22-25. At the same time, the Portuguese have a low level of education, most people come from regionlt10k, and literate people exist only as a very small minority in society. From the box plot of age at marriage and the number of children, we can see that the number of children is inversely proportional to age, and the lower the age of marriage, the higher the number of children. On the contrary, literate women have significantly more children than illiterate women.

```{r}
#| label: tbl-summary
#| tbl-cap: Dataset Summary
#| echo: false
#| warning: false
#| message: false

summary_table <-summary(data)
summary_table 
```

## Model Selection





```{r}
#| include: false
#| warning: false
#| message: false
# Load necessary libraries
library(dplyr)
library(MASS)

# Replace zero or negative values in monthsSinceM with a small positive value (e.g., 0.5)
data$monthsSinceM <- ifelse(data$monthsSinceM <= 0, 0.1, data$monthsSinceM)

# Convert categorical variables
data$ageMarried <- as.factor(data$ageMarried)
data$literacy <- as.factor(data$literacy)

# Rename region variables correctly (Replace dots with underscores)
colnames(data) <- gsub("\\.", "_", colnames(data))

# Fit Poisson regression model with offset (log of monthsSinceM)
poisson_model <- glm(children ~ literacy + ageMarried + region10_20k + 
                     region20k_ + regionlisbon + regionlt10k + regionporto + 
                     offset(log(monthsSinceM)), 
                     family = poisson(link = "log"), data = data)

# Display summary
summary(poisson_model)
```
Our Final model is showing above:
```{=latex}
\begin{align}
\text{children}_i &\sim \text{Poisson}(\lambda_i) \\
\log(\lambda_i) &= -4.13839 
    - 0.12906 \cdot \text{literacy}_i  \\
    &\quad + 0.03536 \cdot \text{ageMarried15to18}_i 
    + 0.02006 \cdot \text{ageMarried18to20}_i \\
    &\quad - 0.01574 \cdot \text{ageMarried20to22}_i 
    - 0.03959 \cdot \text{ageMarried22to25}_i \\
    &\quad - 0.02093 \cdot \text{ageMarried25to30}_i 
    - 0.01054 \cdot \text{ageMarried30toInf}_i \\
    &\quad - 0.06783 \cdot \text{region10\_20k}_i 
    - 0.20067 \cdot \text{region20k\_}_i \\
    &\quad - 0.18386 \cdot \text{regionlisbon}_i 
    + 0.09381 \cdot \text{regionlt10k}_i \\
    &\quad + \text{NA} \cdot \text{regionporto}_i 
    + \log(\text{monthsSinceM}_i)
\end{align}

```
## Model selection (use model with offset)
## Use likelyhood rario test to do the performance ,  to see the efficiency between model
## Test overdispersion

# Conclusion

# Appendix {-}

\newpage

# References {-}
