---
title: "How Literacy and Age at Marriage Shape Family Size in Portugal, 1979"
subtitle: "A Statistic Research Using Poisson Regression Models "
author: 
  - Shanjie Jiao
  - Xizi Sun
thanks: "Code and data are available at: https://github.com/Jie-jiao05/Family_Size."
date: today
date-format: long
abstract: "TBD. IDK. Smash Them"
format:
  pdf:
    fig-pos: H
toc: true
number-sections: true
bibliography: references.bib

---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(dplyr)
library(here)
library(readr)
library(lubridate)
library(rvest)
library(palmerpenguins)
library(maps)
library(ggplot2)
library(knitr)
library(arrow)
library(rstanarm)
library(plotly)
library(patchwork)
library(tidyr)
library(bayesplot)
library(gridExtra)
library(kableExtra)
library(broom.mixed)
library(modelsummary)

data <- read.csv("~/Family_Size/Data/cleaned_portugal.csv")
```

# Introduction
## Why this topic is important
## 3 citation relate with our research 
## Our objective, goal, how the citation related to, what we gonna do
## Brief introduction of our research


# Method
## Use GLM, possion,   (reason), what is x how to interoperate

# Result
## Data Description
This dataset originates from the 1979-80 Portuguese Fertility Survey [@PortugalWFS], conducted as part of the World Fertility Survey (WFS) program and archived by Princeton University’s Office of Population Research (OPR).

The Dataset includes 5,148 observations and 8 variables, covering factors such as literacy, marriage age, pregnancies, children, and regional demographics. @tbl-summary it shows a summary information of our dataset.

To analyze the relationship between family size, literacy, and age at marriage, the variable Literacy was categorized as a dummy variable, assigning values 0 and 1 to illiterate and literate individuals, respectively. The analysis primarily focused on demographic factors, economic influences, and regional differences, aiming to assess their impact on family size. Additionally, an offset variable accounting for the duration of marriage was included to control for varying exposure times.

```{r}
#| label: tbl-summary
#| tbl-cap: Dataset Summary
#| echo: false
#| warning: false
#| message: false

summary_table <-summary(data)
summary_table 
```

## Model Selection





```{r}
#| include: false
#| warning: false
#| message: false
# Load necessary libraries
library(dplyr)
library(MASS)

# Replace zero or negative values in monthsSinceM with a small positive value (e.g., 0.5)
data$monthsSinceM <- ifelse(data$monthsSinceM <= 0, 0.1, data$monthsSinceM)

# Convert categorical variables
data$ageMarried <- as.factor(data$ageMarried)
data$literacy <- as.factor(data$literacy)

# Rename region variables correctly (Replace dots with underscores)
colnames(data) <- gsub("\\.", "_", colnames(data))

# Fit Poisson regression model with offset (log of monthsSinceM)
poisson_model <- glm(children ~ literacy + ageMarried + region10_20k + 
                     region20k_ + regionlisbon + regionlt10k + regionporto + 
                     offset(log(monthsSinceM)), 
                     family = poisson(link = "log"), data = data)

# Display summary
summary(poisson_model)
```
Our Final model is showing above:
```{=latex}
\begin{align}
\text{children}_i &\sim \text{Poisson}(\lambda_i) \\
\log(\lambda_i) &= -4.13839 
    - 0.12906 \cdot \text{literacy}_i  \\
    &\quad + 0.03536 \cdot \text{ageMarried15to18}_i 
    + 0.02006 \cdot \text{ageMarried18to20}_i \\
    &\quad - 0.01574 \cdot \text{ageMarried20to22}_i 
    - 0.03959 \cdot \text{ageMarried22to25}_i \\
    &\quad - 0.02093 \cdot \text{ageMarried25to30}_i 
    - 0.01054 \cdot \text{ageMarried30toInf}_i \\
    &\quad - 0.06783 \cdot \text{region10\_20k}_i 
    - 0.20067 \cdot \text{region20k\_}_i \\
    &\quad - 0.18386 \cdot \text{regionlisbon}_i 
    + 0.09381 \cdot \text{regionlt10k}_i \\
    &\quad + \text{NA} \cdot \text{regionporto}_i 
    + \log(\text{monthsSinceM}_i)
\end{align}

```
## Model selection (use model with offset)
## Use likelyhood rario test to do the performance ,  to see the efficiency between model
## Test overdispersion

# Conclusion

# Appendix {-}

\newpage

# References {-}
