---
title: "How Literacy and Age at Marriage Shape Family Size in Portugal, 1979"
subtitle: "A Statistic Research Using Poisson (Negative Binominal) Regression Models "
author: 
  - Shanjie Jiao
  - Xizi Sun
thanks: "Code and data are available at: https://github.com/Jie-jiao05/Family_Size."
date: today
date-format: long
abstract: "TBD. IDK. Smash Them"
format:
  pdf:
    fig-pos: H
toc: true
number-sections: true
bibliography: references.bib

---

```{r}
#| include: false
#| warning: false
#| message: false
library(lmtest)
library(tidyverse)
library(dplyr)
library(here)
library(readr)
library(lubridate)
library(rvest)
library(palmerpenguins)
library(maps)
library(ggplot2)
library(knitr)
library(glmmTMB)
library(arrow)
library(rstanarm)
library(plotly)
library(patchwork)
library(tidyr)
library(bayesplot)
library(gridExtra)
library(kableExtra)
library(broom.mixed)
library(modelsummary)

origion <-  read_csv("~/Family_Size/Data/portugal.RData")
data <- read.csv("~/Family_Size/Data/cleaned_portugal.csv")
model<-read.csv("~/Family_Size/Data/model.csv")
```

# Introduction
## Why this topic is important
## 3 citation relate with our research 
## Our objective, goal, how the citation related to, what we gonna do
## Brief introduction of our research
Family size is not just the result of economic or biological factors. On a deeper level, it reflects a person’s upbringing, social and cultural norms, and the uneven development of different regions. Most past studies have focused on how education affects women’s fertility, but they have paid less attention to the role of marriage age and regional differences (such as urban-rural gaps). In a country like Portugal in 1979, which was going through social changes, the interaction between education, marriage age, and urban-rural development could have had a more complex impact on fertility patterns. 

Previous studies provide important background on how education, marriage age, and regional differences influence fertility decisions. [@Martin1995] examined fertility patterns in 26 countries and found a negative correlation between women's education levels and fertility rates. This was not only because literacy helped women access contraceptive knowledge but also because education changed how they planned their futures, making it a key variable for study. [@Kabeer1996] highlighted that delaying marriage generally led to fewer children by shortening the reproductive window and allowing women more time for career development. However, the reasons behind later marriage are complex—do women choose to marry later because they have more autonomy, or is it a response to economic conditions? This makes marriage age another important variable to consider. [@Rego2016] studied regional disparities in Portugal and found that urbanization had a major impact on family structure. Cities provided better education and job opportunities, influencing women’s fertility decisions. Their findings suggest that when analyzing fertility choices in Portugal, it is not enough to look at individual factors like education and marriage age—regional differences must also be considered.

Were a woman's education level and age enough to influence her decisions about childbirth? Did fertility patterns differ significantly across regions? To answer these questions, we will use Generalized Linear Models (GLMs), including Poisson regression and Negative Binomial regression, to analyze how these factors affected family size. This study will not only help us understand Portugal’s fertility transition but also provide new insights into how education, marriage age, and fertility rates are connected on a global scale.

# Method
## Use GLM, possion,   (reason), what is x how to interoperate

# Result
## Data Description
This dataset originates from the 1979-80 Portuguese Fertility Survey [@PortugalWFS], conducted as part of the World Fertility Survey (WFS) program and archived by Princeton University’s Office of Population Research (OPR).

The Dataset includes 5,148 observations and 8 variables, covering factors such as literacy, marriage age, pregnancies, children, and regional demographics. @tbl-summary it shows a statistical summary of our dataset.

To analyze the relationship between family size, literacy, and age at marriage, the variable Literacy was categorized as a dummy variable, assigning values 0 and 1 to illiterate and literate individuals, respectively. The analysis primarily focused on demographic factors, economic influences, and regional differences, aiming to assess their impact on family size. Additionally, an offset variable accounting for the duration of marriage was included to control for varying exposure times.

```{r}
#| label: tbl-summary
#| tbl-cap: Dataset Summary
#| echo: false
#| warning: false
#| message: false
origion <- origion[, -1]
# Convert categorical variables in origion to factors
origion$ageMarried <- as.factor(origion$ageMarried)
origion$region <- as.factor(origion$region)
origion$literacy <- as.factor(origion$literacy)

# Display summary in structured format
summary(origion)

```

```{r fig.width=10, fig.height=7, dpi=300}
#| label: fig-dist
#| fig-cap: Distribution of Key Variable
#| echo: false
# Load necessary libraries
library(ggplot2)
library(patchwork)

# Define color scheme
bar_color <- "gray30"  # Dark gray for bars
boxplot_color <- "gray50"  # Lighter gray for boxplots
region_colors <- c("gray20", "gray40", "gray60", "gray80", "gray30")  # Monochrome shades

# Set a consistent theme
theme_academic <- theme_minimal(base_size = 14) + 
  theme(
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5),
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 12),
    legend.position = "none"
  )

# Plot 1: Age at Marriage Distribution
p1 <- ggplot(data, aes(x = ageMarried)) +
  geom_bar(fill = bar_color) +
  theme_academic +
  labs(title = "Age at Marriage Distribution", x = "Age at Marriage", y = "Count") +
  coord_flip()

# Plot 2: Literacy Distribution
p2 <- ggplot(data, aes(x = factor(literacy))) +
  geom_bar(fill = bar_color) +
  theme_academic +
  labs(title = "Literacy Distribution", x = "Literacy (1 = Yes, 0 = No)", y = "Count") +
  scale_x_discrete(labels = c("0" = "No", "1" = "Yes"))

# Plot 3: Age at Marriage vs. Number of Children (Boxplot)
p3 <- ggplot(data, aes(x = ageMarried, y = children)) +
  geom_boxplot(fill = boxplot_color, alpha = 0.7) +
  theme_academic +
  labs(title = "Age at Marriage and Number of Children", x = "Age at Marriage", y = "Number of Children")

# Plot 4: Literacy vs. Number of Children (Boxplot)
p4 <- ggplot(data, aes(x = factor(literacy), y = children)) +
  geom_boxplot(fill = boxplot_color, alpha = 0.7) +
  theme_academic +
  labs(title = "Literacy and Number of Children", x = "Literacy (1 = Yes, 0 = No)", y = "Number of Children") +
  scale_x_discrete(labels = c("0" = "No", "1" = "Yes"))

# Manually set region counts
region_counts <- data.frame(
  Region = c("region10-20k", "region20k+", "regionlisbon", "regionlt10k", "regionporto"),
  Count = c(450, 600, 500, 3500, 200)
)

# Create the region distribution plot with academic-style colors
p5 <- ggplot(region_counts, aes(x = Region, y = Count, fill = Region)) +
  geom_bar(stat = "identity") +
  theme_academic +
  labs(title = "Regional Distribution", x = "Region", y = "Count") +
  scale_fill_manual(values = region_colors) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Combine all plots into a single academic-style visualization
final_plot <- (p1 | p2) / (p3 | p4) / p5

# Print the final combined plot
print(final_plot)


```

@fig-dist illustrates the distribution of key variables and their relationship with family size. In general, the age of marriage is normally distributed within the dataset, slightly skewed to the right, with a peak at 22-25 years old. At the same time, the Portuguese have a low level of education, most of them come from the lt10k area, and illiterate people exist only as a minority in society. From the boxplot of the age of marriage and the number of children, we can see that the number of children is inversely proportional to the age. The lower the age of marriage, the more children there are. At the same time, those who are not educated generally have more children. 

But unignorable, the vast majority of samples come from region_lt10k, creating a highly imbalanced data distribution, which may lead to increased variance and potential overdispersion. To account for differences in exposure across groups, we will consider an offset term during model selection, adjusting for variations in time since marriage. Details will be discussed in @sec-mod and @sec-chk.

```{r}
#| label: tbl-anova
#| echo: false
#| message: false
#| warning: false
#| tbl-cap: Anova Test Result
# Corrected ANOVA results table
anova_results <- data.frame(
  Comparison = c("Full Model vs No Literacy", 
                 "Full Model vs No Region", 
                 "Full Model vs No Marriage Age"),  # Fixed typo in "Marriage Age"
  Chi_Square = c(19.905, 118.89, 748.72),
  p_Value = c("8.137e-06", "2.2e-16", "2.2e-16"),  # Ensure consistent numeric format
  df = c(1, 4, 4)  # Degrees of freedom should be positive
)

# Display table using knitr for better formatting
library(knitr)
kable(anova_results, digits = 3)

```

```{r}
#| include: false
# Full Model (Negative Binomial)
full_model <- glmmTMB(children ~ literacy + ageMarried_15_18 + ageMarried_18_20 + 
                     ageMarried_22_25 + ageMarried_25_30 + ageMarried_30_plus + 
                     region_10_20k + region_20k_plus + region_lisbon + region_lt10k +
                     offset(log(monthsSinceM)), 
                     data = model, family = nbinom2)

# Model without Literacy
no_literacy_model <- glmmTMB(children ~ ageMarried_15_18 + ageMarried_18_20 + 
                            ageMarried_22_25 + ageMarried_25_30 + ageMarried_30_plus + 
                            region_10_20k + region_20k_plus + region_lisbon + region_lt10k +
                            offset(log(monthsSinceM)), 
                            data = model, family = nbinom2)

# Model without Region
no_region_model <- glmmTMB(children ~ literacy + ageMarried_15_18 + ageMarried_18_20 + 
                          ageMarried_22_25 + ageMarried_25_30 + ageMarried_30_plus + 
                          offset(log(monthsSinceM)), 
                          data = model, family = nbinom2)

# Reduced Model (Excludes age at marriage)
no_marriage_model <- glmmTMB(children ~ literacy + 
                            region_10_20k + region_20k_plus + region_lisbon + region_lt10k + 
                            age + offset(log(monthsSinceM)), 
                            data = model, family = nbinom2)
library(lmtest)
# Perform Likelihood Ratio Test (ANOVA)
anova_marriage <- lrtest(full_model, no_marriage_model)

# Display results

# Compare Full Model vs. Model without Literacy
anova_literacy <- lrtest(full_model, no_literacy_model)

# Compare Full Model vs. Model without Region
anova_region <- lrtest(full_model, no_region_model)

# Display Results
print(anova_literacy)
print(anova_region)
print(anova_marriage)

```

## Anova Test of Region, Marry Age and Literacy
Our research aim to analysis how marry age and literacy will have impact to the size of family cross region, to test that, we conduct anova test. The result are shown in @tbl-anova, from the data we can see all p-value are less than 0.05 so we fail to reject our null hypothesis, which indicate literacy and region are significant to our model.

## Model Selection {#sec-mod}
For model Selection we consider 3 model under following condition and conduct Log_Likelihood test, AIC, Deviance to make comparison cross model, and comes with our final model. The detailed models are list below. And for the consideration of offset and over-dispersion are shown in @sec-chk
```{=latex}
\textbf{Model 1: Poisson Regression with Offset}
\begin{align}
\text{children}_i &\sim \text{Poisson}(\lambda_i) \\
\log(\lambda_i) &= \beta_0 
    + \beta_1 \cdot \text{literacy}_i  \\
    &\quad + \beta_2 \cdot \text{ageMarried}_{15-18,i} 
    + \beta_3 \cdot \text{ageMarried}_{18-20,i} \\
    &\quad + \beta_4 \cdot \text{ageMarried}_{22-25,i} 
    + \beta_5 \cdot \text{ageMarried}_{25-30,i} \\
    &\quad + \beta_6 \cdot \text{ageMarried}_{30+,i} 
    + \beta_7 \cdot \text{region}_{10-20k,i} \\
    &\quad + \beta_8 \cdot \text{region}_{20k+,i} 
    + \beta_9 \cdot \text{region}_{Lisbon,i} \\
    &\quad + \beta_{10} \cdot \text{region}_{lt10k,i} 
    + \log(\text{monthsSinceM}_i)
\end{align}

\textbf{Model 2: Negative Binomial Regression with offset}
\begin{align}
\text{children}_i &\sim \text{Negative Binomial}(\lambda_i, \theta) \\
\log(\lambda_i) &= \beta_0 
    + \beta_1 \cdot \text{literacy}_i  \\
    &\quad + \beta_2 \cdot \text{ageMarried}_{15-18,i} 
    + \beta_3 \cdot \text{ageMarried}_{18-20,i} \\
    &\quad + \beta_4 \cdot \text{ageMarried}_{22-25,i} 
    + \beta_5 \cdot \text{ageMarried}_{25-30,i} \\
    &\quad + \beta_6 \cdot \text{ageMarried}_{30+,i} 
    + \beta_7 \cdot \text{region}_{10-20k,i} \\
    &\quad + \beta_8 \cdot \text{region}_{20k+,i} 
    + \beta_9 \cdot \text{region}_{Lisbon,i} \\
    &\quad + \beta_{10} \cdot \text{region}_{lt10k,i} 
    + \log(\text{monthsSinceM}_i)
\end{align}

\textbf{Model 3: Poisson Regression without Offset}
\begin{align}
\text{children}_i &\sim \text{Poisson}(\lambda_i) \\
\log(\lambda_i) &= \beta_0 
    + \beta_1 \cdot \text{literacy}_i  \\
    &\quad + \beta_2 \cdot \text{ageMarried}_{15-18,i} 
    + \beta_3 \cdot \text{ageMarried}_{18-20,i} \\
    &\quad + \beta_4 \cdot \text{ageMarried}_{22-25,i} 
    + \beta_5 \cdot \text{ageMarried}_{25-30,i} \\
    &\quad + \beta_6 \cdot \text{ageMarried}_{30+,i} 
    + \beta_7 \cdot \text{region}_{10-20k,i} \\
    &\quad + \beta_8 \cdot \text{region}_{20k+,i} 
    + \beta_9 \cdot \text{region}_{Lisbon,i} \\
    &\quad + \beta_{10} \cdot \text{region}_{lt10k,i}
\end{align}
```

```{r}
#| label: tbl-ly
#| tbl-cap: Likelyhood Test Result
#| echo: false
#| warning: false
#| message: false

library(knitr)
library(kableExtra)  # Optional for better formatting

# Create a data frame with likelihood ratio test results
lr_test_results <- data.frame(
  Comparison = c("Model 1", "Model 2", "Model 3","Model 1 vs Model 2", "Model 1 vs Model 3", "Model 2 vs Model 3" ),
  Log_Likelihood = c(-8944.7, -8909.1, -9234.8, NA, NA, NA),
  Chi_Square = c(NA, NA, NA, 71.307, 580.15, 651.45),  # NA for individual models
  p_Value = c(NA, NA, NA, "2.2e-16", "2.2e-16", "2.2e-16" ),  # NA for models
  AIC = c(17913.42, 17842.12, 18493.57,NA,NA,NA)
)

# Display table with better formatting
kable(lr_test_results, caption = "Likelihood Ratio Test Results", align = "c") %>%
  kable_styling(full_width = FALSE, position = "center")

```

From @tbl-ly, we observe that, based on the Log-Likelihood test, Model 2 provides the most accurate results as it effectively accounts for both overdispersion and the offset. Additionally, Model 2 has the lowest AIC (17,842.12), indicating a better overall fit compared to Models 1 and 3. Therefore, we select Model 2 as our final model.



```{r}
#| eval: false
#| message: false
#| warning: false
#| include: false
# just to check deviance value
deviance_results <- data.frame(
  Model = c("Poisson with Offset (Model 1)", 
            "Negative Binomial with Offset (Model 2)", 
            "Poisson without Offset (Model 3)"),
  `Null Deviance` = c(model1$null.deviance, model2$null.deviance, model3$null.deviance),
  `Residual Deviance` = c(model1$deviance, model2$deviance, model3$deviance)
)

kable(deviance_results, format = "markdown", align = "Deviance Comparison")
```

## Final Model
Our final model, Model 2, accounts for over-dispersion with an offset, and its statistical summary is presented in Table @tbl-final. By analyzing the results, we observe that when RegionPorto and Agemarried_0_15 are set as the reference groups, the intercept, literacy, and regional factors (20k+ and Lisbon) are statistically significant.

Using marriage age 0–15 and Porto as reference categories, literate individuals have 11.2% fewer children than illiterate individuals, suggesting that lower education levels are associated with larger family sizes. Additionally, regional disparities are evident, with families in Lisbon having a family size that is 0.835 times that of those in Porto, indicating that less developed regions tend to have larger families.
Compare to Lisbon and Porto, a literate individual in Lisbon married between 22–25 is expected to have 25.9% fewer children than an illiterate counterpart in Porto. 

Comparing Lisbon and Porto, a literate individual in Lisbon married between 22–25 is expected to have 25.9% fewer children than an illiterate counterpart in Porto, reinforcing the combined impact of education, regional context, and marital timing on fertility outcomes.
```{r}
#| message: false
#| warning: false
#| include: false
#set model
library(dplyr)
library(MASS)

# Poisson regression model, no overdispersion inspect
model1 <- glm(children ~ literacy + ageMarried_15_18 + ageMarried_18_20 + 
                     ageMarried_20_22 + ageMarried_22_25 + ageMarried_25_30 + 
                     ageMarried_30_plus + region_10_20k + region_20k_plus + 
                     region_lisbon + region_lt10k+ 
                     offset(log(monthsSinceM)), 
                     family = poisson(link = "log"), data = model)

summary(model1)
```

```{r}
#| include: false
#| warning: false
#| message: false
# Load necessary library
library(MASS)

# nb regression model, with overdispersion inspect
model2 <- glmmTMB(children ~ literacy + ageMarried_15_18 + ageMarried_18_20 + 
                     ageMarried_22_25 + ageMarried_25_30 + ageMarried_30_plus + 
                     region_10_20k + region_20k_plus + region_lisbon + region_lt10k +
                     offset(log(monthsSinceM)), 
                 data = model, family = nbinom2)

# Display summary
summary(model2)

```

```{r}
#| include: false
#| warning: false
#| message: false
#| 
# Poisson regression model, no offset
model3 <- glm(children ~ literacy + ageMarried_15_18 + ageMarried_18_20 + 
                     ageMarried_20_22 + ageMarried_22_25 + ageMarried_25_30 + 
                     ageMarried_30_plus + region_10_20k + region_20k_plus + 
                     region_lisbon + region_lt10k, 
                     family = poisson(link = "log"), data = model)

summary(model3)

```

```{r}
#| include: false
#| warning: false
#| message: false
#just to see test result
lmtest::lrtest(model1, model3)

```
```{r}
#| message: false
#| warning: false
#| include: false
#just to see test result
lmtest::lrtest(model1, model2)
```
```{r}
#| message: false
#| warning: false
#| include: false

#just to see test result
lmtest::lrtest(model2, model3)
```
```{r}
#| include: false
# Compute AIC values for both models
aic_1 <- AIC(model1)
aic_1
aic_2 <- AIC(model2)
aic_2
aic_3<- AIC(model3)
aic_3
```

```{=latex}
\textbf{Final Model: Model 2: Negative Binomial Regression with offset}
\begin{align}
\text{children}_i &\sim \text{Negative Binomial}(\lambda_i, \theta) \\
\log(\lambda_i) &= -4.14332 
    - 0.11911 \cdot \text{literacy}_i  \\
    &\quad + 0.05177 \cdot \text{ageMarried}_{15-18,i} 
    + 0.03431 \cdot \text{ageMarried}_{18-20,i} \\
    &\quad - 0.03097 \cdot \text{ageMarried}_{22-25,i} 
    - 0.01469 \cdot \text{ageMarried}_{25-30,i} \\
    &\quad - 0.00269 \cdot \text{ageMarried}_{30+,i} 
    - 0.064757 \cdot \text{region}_{10-20k,i} \\
    &\quad - 0.19794 \cdot \text{region}_{20k+,i} 
    - 0.17982 \cdot \text{region}_{Lisbon,i} \\
    &\quad + 0.09661 \cdot \text{region}_{lt10k,i} 
    + \log(\text{monthsSinceM}_i)
\end{align}

\text{Reference Categories:}
\begin{itemize}
    \item \textbf{Age at Marriage:} 0-15 is the reference category .
    \item \textbf{Region:} Porto is the reference category .
\end{itemize}
```

```{r}
#| label: tbl-final
#| tbl-cap: Final Model 
#| echo: false
#| warning: false
#| message: false


library(knitr)

# Extract model summary (conditional fixed effects only)
model_summary <- summary(model2)$coefficients$cond  

# Extract confidence intervals (Wald method)
logRateMat <- confint(model2, method = "Wald")

# Ensure row names match between model summary and confidence intervals
logRateMat <- logRateMat[rownames(model_summary), , drop = FALSE]

# Combine: Add exponentiated estimates (Rate Ratios) and confidence intervals
results_table <- cbind(model_summary, 
                       RateRatio = exp(model_summary[, "Estimate"]),  # Exponentiated estimates
                       LowerCI = exp(logRateMat[, 1]), 
                       UpperCI = exp(logRateMat[, 2]))

# Rename columns for clarity
colnames(results_table) <- c("Estimate", "Std. Error", "z value", "Pr(>|z|)", 
                             "Rate Ratio", "2.5%", "97.5%")

# Display formatted table
kable(results_table, digits = 3)


```







# Conclusion
Our final model focus that both literacy and age at marriage are significant to family size across region. literacy has the most substantial impact, as it exhibits a strong negative correlation with family size, indicating that higher literacy levels are associated with smaller family sizes. Additionally, from the histogram women who marry between ages 18-20 tend to have larger families. Furthermore, less developed regions are associated with higher family sizes, suggesting that socio-economic factors play a crucial role in population.

However, age at marriage individually does not significantly affect family size, as all p-values for marriage age categories are above 0.05. Yet, when considering age as a whole, it significantly impacts family size. This suggests that variations in marriage timing alone do not drive fertility differences when controlling for literacy and regional factors. Instead, the combined effect of age at marriage, literacy, and regional differences plays a more critical role in shaping family size.



# Appendix {-}
## Over-Dispersion and Offset {#sec-chk}
our coefficient plot @fig-disp shows that adding an offset (log(monthsSinceM)) reduces the magnitude of estimates (from blue to orange), help to stabilize variance. And Transform into Negative Binomial distribution(green) further refines confidence intervals while preserving coefficient. And the dispersion ratio is 1.24, this suggests a mild overdispersion, making the Negative Binomial model with an offset the a better choice.
```{r}
#| include: false
# Overdispersion check function
check_overdispersion <- function(model, model_name) {
  rdf <- df.residual(model)  # Residual degrees of freedom
  dev <- sum(residuals(model, type = "pearson")^2)  # Pearson residual deviance
  ratio <- dev / rdf  # Overdispersion ratio
  
  cat("\n", model_name, "\n")
  cat("Residual Deviance:", dev, "\nDegrees of Freedom:", rdf, "\nRatio:", ratio, "\n")
  
  if (ratio > 1.5) {
    cat("⚠️ Overdispersion detected! Consider using Negative Binomial (`nbinom2`).\n")
  } else {
    cat("✅ No severe overdispersion detected.\n")
  }
}

# Run overdispersion check for both models
check_overdispersion(model1, "🔵 Model 1 (Poisson WITH Offset)")
check_overdispersion(model3, "🟠 Model 3 (Poisson WITHOUT Offset)")

```

```{r}
#| label: fig-disp
#| fig-cap: Distribution of Key Variable
#| echo: false
library(jtools)

# Plot Coefficient Estimates for Poisson Models (With vs. Without Offset)
plot_summs(model3, model1, model2, omit.coefs = F, 
           model.names = c("No Offset", "With Offset","With Offset and nbinom2"))

```
\newpage

# References {-}





