---
title: "How Literacy and Age at Marriage Shape Family Size in Portugal, 1979"
subtitle: "A Statistic Research Using Poisson (Negative Binominal) Regression Models "
author: 
  - Shanjie Jiao
  - Xizi Sun
thanks: "Code and data are available at: https://github.com/Jie-jiao05/Family_Size."
date: today
date-format: long
abstract: "TBD. IDK. Smash Them"
format:
  pdf:
    fig-pos: H
toc: true
number-sections: true
bibliography: references.bib

---

```{r}
#| include: false
#| warning: false
#| message: false
library(lmtest)
library(tidyverse)
library(dplyr)
library(here)
library(readr)
library(lubridate)
library(rvest)
library(palmerpenguins)
library(maps)
library(ggplot2)
library(knitr)
library(arrow)
library(rstanarm)
library(plotly)
library(patchwork)
library(tidyr)
library(bayesplot)
library(gridExtra)
library(kableExtra)
library(broom.mixed)
library(modelsummary)

origion <-  read_csv("~/Family_Size/Data/portugal.RData")
data <- read.csv("~/Family_Size/Data/cleaned_portugal.csv")
model<-read.csv("~/Family_Size/Data/model.csv")
```

# Introduction
## Why this topic is important
## 3 citation relate with our research 
## Our objective, goal, how the citation related to, what we gonna do
## Brief introduction of our research


# Method
## Use GLM, possion,   (reason), what is x how to interoperate

# Result
## Data Description
This dataset originates from the 1979-80 Portuguese Fertility Survey [@PortugalWFS], conducted as part of the World Fertility Survey (WFS) program and archived by Princeton University’s Office of Population Research (OPR).

The Dataset includes 5,148 observations and 8 variables, covering factors such as literacy, marriage age, pregnancies, children, and regional demographics. @tbl-summary it shows a statistical summary of our dataset.

To analyze the relationship between family size, literacy, and age at marriage, the variable Literacy was categorized as a dummy variable, assigning values 0 and 1 to illiterate and literate individuals, respectively. The analysis primarily focused on demographic factors, economic influences, and regional differences, aiming to assess their impact on family size. Additionally, an offset variable accounting for the duration of marriage was included to control for varying exposure times.

```{r}
#| label: tbl-summary
#| tbl-cap: Dataset Summary
#| echo: false
#| warning: false
#| message: false
origion <- origion[, -1]
# Convert categorical variables in origion to factors
origion$ageMarried <- as.factor(origion$ageMarried)
origion$region <- as.factor(origion$region)
origion$literacy <- as.factor(origion$literacy)

# Display summary in structured format
summary(origion)

```

```{r fig.width=10, fig.height=7, dpi=300}
#| label: fig-dist
#| fig-cap: Distribution of Key Variable
#| echo: false
# Load necessary libraries
library(ggplot2)
library(patchwork)

# Define color scheme
bar_color <- "gray30"  # Dark gray for bars
boxplot_color <- "gray50"  # Lighter gray for boxplots
region_colors <- c("gray20", "gray40", "gray60", "gray80", "gray30")  # Monochrome shades

# Set a consistent theme
theme_academic <- theme_minimal(base_size = 14) + 
  theme(
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5),
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 12),
    legend.position = "none"
  )

# Plot 1: Age at Marriage Distribution
p1 <- ggplot(data, aes(x = ageMarried)) +
  geom_bar(fill = bar_color) +
  theme_academic +
  labs(title = "Age at Marriage Distribution", x = "Age at Marriage", y = "Count") +
  coord_flip()

# Plot 2: Literacy Distribution
p2 <- ggplot(data, aes(x = factor(literacy))) +
  geom_bar(fill = bar_color) +
  theme_academic +
  labs(title = "Literacy Distribution", x = "Literacy (1 = Yes, 0 = No)", y = "Count") +
  scale_x_discrete(labels = c("0" = "No", "1" = "Yes"))

# Plot 3: Age at Marriage vs. Number of Children (Boxplot)
p3 <- ggplot(data, aes(x = ageMarried, y = children)) +
  geom_boxplot(fill = boxplot_color, alpha = 0.7) +
  theme_academic +
  labs(title = "Age at Marriage and Number of Children", x = "Age at Marriage", y = "Number of Children")

# Plot 4: Literacy vs. Number of Children (Boxplot)
p4 <- ggplot(data, aes(x = factor(literacy), y = children)) +
  geom_boxplot(fill = boxplot_color, alpha = 0.7) +
  theme_academic +
  labs(title = "Literacy and Number of Children", x = "Literacy (1 = Yes, 0 = No)", y = "Number of Children") +
  scale_x_discrete(labels = c("0" = "No", "1" = "Yes"))

# Manually set region counts
region_counts <- data.frame(
  Region = c("region10-20k", "region20k+", "regionlisbon", "regionlt10k", "regionporto"),
  Count = c(450, 600, 500, 3500, 200)
)

# Create the region distribution plot with academic-style colors
p5 <- ggplot(region_counts, aes(x = Region, y = Count, fill = Region)) +
  geom_bar(stat = "identity") +
  theme_academic +
  labs(title = "Regional Distribution", x = "Region", y = "Count") +
  scale_fill_manual(values = region_colors) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Combine all plots into a single academic-style visualization
final_plot <- (p1 | p2) / (p3 | p4) / p5

# Print the final combined plot
print(final_plot)


```
@fig-dist illustrates the distribution of key variables and their relationship with family size. In general, the age of marriage is normally distributed within the dataset, slightly skewed to the right, with a peak at 22-25 years old. At the same time, the Portuguese have a low level of education, most of them come from the lt10k area, and illiterate people exist only as a minority in society. From the boxplot of the age of marriage and the number of children, we can see that the number of children is inversely proportional to the age. The lower the age of marriage, the more children there are. At the same time, those who are not educated generally have more children.

## Model Selection
For model Selection we consider 3 model under following condition and conduct Log_Likelihood test, AIC, Deviance to make comparision cross model, and comes with our final model.
```{=latex}
\textbf{Model 1: Poisson Regression with Offset}
\begin{align}
\text{children}_i &\sim \text{Poisson}(\lambda_i) \\
\log(\lambda_i) &= \beta_0 
    + \beta_1 \cdot \text{literacy}_i  \\
    &\quad + \beta_2 \cdot \text{ageMarried}_{15-18,i} 
    + \beta_3 \cdot \text{ageMarried}_{18-20,i} \\
    &\quad + \beta_4 \cdot \text{ageMarried}_{22-25,i} 
    + \beta_5 \cdot \text{ageMarried}_{25-30,i} \\
    &\quad + \beta_6 \cdot \text{ageMarried}_{30+,i} 
    + \beta_7 \cdot \text{region}_{10-20k,i} \\
    &\quad + \beta_8 \cdot \text{region}_{20k+,i} 
    + \beta_9 \cdot \text{region}_{Lisbon,i} \\
    &\quad + \beta_{10} \cdot \text{region}_{lt10k,i} 
    + \log(\text{monthsSinceM}_i)
\end{align}

\textbf{Model 2: Negative Binomial Regression with offset}
\begin{align}
\text{children}_i &\sim \text{Negative Binomial}(\lambda_i, \theta) \\
\log(\lambda_i) &= \beta_0 
    + \beta_1 \cdot \text{literacy}_i  \\
    &\quad + \beta_2 \cdot \text{ageMarried}_{15-18,i} 
    + \beta_3 \cdot \text{ageMarried}_{18-20,i} \\
    &\quad + \beta_4 \cdot \text{ageMarried}_{22-25,i} 
    + \beta_5 \cdot \text{ageMarried}_{25-30,i} \\
    &\quad + \beta_6 \cdot \text{ageMarried}_{30+,i} 
    + \beta_7 \cdot \text{region}_{10-20k,i} \\
    &\quad + \beta_8 \cdot \text{region}_{20k+,i} 
    + \beta_9 \cdot \text{region}_{Lisbon,i} \\
    &\quad + \beta_{10} \cdot \text{region}_{lt10k,i} 
    + \log(\text{monthsSinceM}_i)
\end{align}

\textbf{Model 3: Poisson Regression without Offset}
\begin{align}
\text{children}_i &\sim \text{Poisson}(\lambda_i) \\
\log(\lambda_i) &= \beta_0 
    + \beta_1 \cdot \text{literacy}_i  \\
    &\quad + \beta_2 \cdot \text{ageMarried}_{15-18,i} 
    + \beta_3 \cdot \text{ageMarried}_{18-20,i} \\
    &\quad + \beta_4 \cdot \text{ageMarried}_{22-25,i} 
    + \beta_5 \cdot \text{ageMarried}_{25-30,i} \\
    &\quad + \beta_6 \cdot \text{ageMarried}_{30+,i} 
    + \beta_7 \cdot \text{region}_{10-20k,i} \\
    &\quad + \beta_8 \cdot \text{region}_{20k+,i} 
    + \beta_9 \cdot \text{region}_{Lisbon,i} \\
    &\quad + \beta_{10} \cdot \text{region}_{lt10k,i}
\end{align}
```

```{r}
#| label: tbl-ly
#| tbl-cap: Likelyhood Test Result
#| echo: false
#| warning: false
#| message: false

library(knitr)
library(kableExtra)  # Optional for better formatting

# Create a data frame with likelihood ratio test results
lr_test_results <- data.frame(
  Comparison = c("Model 1 vs Model 2", "Model 1 vs Model 3", "Model 2 vs Model 3", 
                 "Model 1", "Model 2", "Model 3"),
  Log_Likelihood = c(-8944.7, -9234.8, -8909.1, -NA, -NA, -NA),
  Chi_Square = c(71.307, 580.15, 651.45, NA, NA, NA),  # NA for individual models
  p_Value = c("2.2e-16", "2.2e-16", "2.2e-16", NA, NA, NA),  # NA for models
  Deviance = c(NA, NA, NA, 5654.313, 4912.803, 6936.059),  # Deviance only for models
  AIC = c(NA,NA,NA,17913.42, 17842.12, 18493.57 )
)

# Display table with better formatting
kable(lr_test_results, caption = "Likelihood Ratio Test Results", align = "c") %>%
  kable_styling(full_width = FALSE, position = "center")

```

From @tbl-ly, we observe that, based on the Log-Likelihood test, Model 2 provides the most accurate results as it effectively accounts for both overdispersion and the offset. Additionally, Model 2 has the lowest AIC (17,842.12) and deviance (4,912.803), indicating a better overall fit compared to Models 1 and 3. Therefore, we select Model 2 as our final model.



```{r}
#| eval: false
#| message: false
#| warning: false
#| include: false
# just to check deviance value
deviance_results <- data.frame(
  Model = c("Poisson with Offset (Model 1)", 
            "Negative Binomial with Offset (Model 2)", 
            "Poisson without Offset (Model 3)"),
  `Null Deviance` = c(model1$null.deviance, model2$null.deviance, model3$null.deviance),
  `Residual Deviance` = c(model1$deviance, model2$deviance, model3$deviance)
)

kable(deviance_results, format = "markdown", align = "Deviance Comparison")
```

## Final Model
Our final model:model 2 consider over-dispersion with offset is showing below, statistic summary are showing in @tbl-final

```{r}
#| message: false
#| warning: false
#| include: false
#set model
library(dplyr)
library(MASS)

# Poisson regression model, no overdispersion inspect
model1 <- glm(children ~ literacy + ageMarried_15_18 + ageMarried_18_20 + 
                     ageMarried_20_22 + ageMarried_22_25 + ageMarried_25_30 + 
                     ageMarried_30_plus + region_10_20k + region_20k_plus + 
                     region_lisbon + region_lt10k+ 
                     offset(log(monthsSinceM)), 
                     family = poisson(link = "log"), data = model)

summary(model1)
```

```{r}
#| include: false
#| warning: false
#| message: false
# Load necessary library
library(MASS)

# nb regression model, with overdispersion inspect
model2 <- glm.nb(children ~ literacy + ageMarried_15_18 + ageMarried_18_20 + 
                     ageMarried_22_25 + ageMarried_25_30 + ageMarried_30_plus + 
                     region_10_20k + region_20k_plus + region_lisbon + region_lt10k +
                     offset(log(monthsSinceM)), 
                 data = model)

# Display summary
summary(model2)

```

```{r}
#| include: false
#| warning: false
#| message: false
#| 
# Poisson regression model, no offset
model3 <- glm(children ~ literacy + ageMarried_15_18 + ageMarried_18_20 + 
                     ageMarried_20_22 + ageMarried_22_25 + ageMarried_25_30 + 
                     ageMarried_30_plus + region_10_20k + region_20k_plus + 
                     region_lisbon + region_lt10k, 
                     family = poisson(link = "log"), data = model)

summary(model3)

```

```{r}
#| include: false
#| warning: false
#| message: false
#just to see test result
lmtest::lrtest(model1, model3)

```
```{r}
#| message: false
#| warning: false
#| include: false
#just to see test result
lmtest::lrtest(model1, model2)
```
```{r}
#| message: false
#| warning: false
#| include: false

#just to see test result
lmtest::lrtest(model2, model3)
```
```{r}
#| include: false
# Compute AIC values for both models
aic_1 <- AIC(model1)
aic_1
aic_2 <- AIC(model2)
aic_2
aic_3<- AIC(model3)
aic_3
```

```{=latex}
\textbf{Final Model: Model 2: Negative Binomial Regression with offset}
\begin{align}
\text{children}_i &\sim \text{Negative Binomial}(\lambda_i, \theta) \\
\log(\lambda_i) &= -4.143302 
    - 0.119106 \cdot \text{literacy}_i  \\
    &\quad + 0.051760 \cdot \text{ageMarried}_{15-18,i} 
    + 0.034304 \cdot \text{ageMarried}_{18-20,i} \\
    &\quad - 0.030966 \cdot \text{ageMarried}_{22-25,i} 
    - 0.014696 \cdot \text{ageMarried}_{25-30,i} \\
    &\quad - 0.002713 \cdot \text{ageMarried}_{30+,i} 
    - 0.064757 \cdot \text{region}_{10-20k,i} \\
    &\quad - 0.197944 \cdot \text{region}_{20k+,i} 
    - 0.179831 \cdot \text{region}_{Lisbon,i} \\
    &\quad + 0.096595 \cdot \text{region}_{lt10k,i} 
    + \log(\text{monthsSinceM}_i)
\end{align}

\text{Reference Categories:}
\begin{itemize}
    \item \textbf{Age at Marriage:} 0-15 is the reference category .
    \item \textbf{Region:} Porto is the reference category .
\end{itemize}
```

```{r}
#| label: tbl-final
#| tbl-cap: Final Model 
#| echo: false
#| warning: false
#| message: false
# Extract model summary with confidence intervals
logRateMat <- confint(model2)  # Extract confidence intervals
model_summary <- summary(model2)$coefficients  # Extract coefficients

# Combine coefficients, exponentiated estimates (est), and confidence intervals
results_table <- cbind(model_summary, exp(model_summary[,1]), exp(logRateMat))

# Rename columns for clarity
colnames(results_table) <- c("Estimate", "Std. Error", "z value", "Pr(>|z|)", "est", "Lower CI", "Upper CI")

# Display formatted table in R console
knitr::kable(results_table, digits = 3)
```
## Anova Test of Region, Marry Age and Literacy
Our research aim to analysis how marry age and literacy will have impact to the size of family cross region, to test that, we conduct anova test. The result are shown in @tbl-anova, from the data we can see but p-value are less than 0.05 so we fail to reject our null hypothesis, which indicate literacy and region are significant to our model.
```{r}
#| label: tbl-anova
#| echo: false
#| message: false
#| warning: false
#| tbl-cap: Anova Test Result
# Load necessary package
library(knitr)

# Create a data frame for ANOVA test results
anova_results <- data.frame(
  Comparison = c("Full Model vs No Literacy", "Full Model vs No Region","Full Model vs No Marrage Age"),
  Chi_Square = c(19.905, 118.89, 748.72),
  p_Value = c(8.137e-06, "< 2.2e-16", "< 2.2e-16 ")

)

# Display the table in a formatted way
kable(anova_results)

```

```{r}
#| include: false
# Full Model (Negative Binomial)
full_model <- glm.nb(children ~ literacy + ageMarried_15_18 + ageMarried_18_20 + 
                     ageMarried_22_25 + ageMarried_25_30 + ageMarried_30_plus + 
                     region_10_20k + region_20k_plus + region_lisbon + region_lt10k +
                     offset(log(monthsSinceM)), 
                     data = model)

# Model without Literacy
no_literacy_model <- glm.nb(children ~ ageMarried_15_18 + ageMarried_18_20 + 
                            ageMarried_22_25 + ageMarried_25_30 + ageMarried_30_plus + 
                            region_10_20k + region_20k_plus + region_lisbon + region_lt10k +
                            offset(log(monthsSinceM)), 
                            data = model)

# Model without Region
no_region_model <- glm.nb(children ~ literacy + ageMarried_15_18 + ageMarried_18_20 + 
                          ageMarried_22_25 + ageMarried_25_30 + ageMarried_30_plus + 
                          offset(log(monthsSinceM)), 
                          data = model)

# Reduced Model (Excludes age at marriage)
no_marriage_model <- glm.nb(children ~ literacy + 
                            region_10_20k + region_20k_plus + region_lisbon + region_lt10k + 
                            age + offset(log(monthsSinceM)), 
                            data = model)

# Perform Likelihood Ratio Test (ANOVA)
anova_marriage <- lrtest(full_model, no_marriage_model)

# Display results

# Compare Full Model vs. Model without Literacy
anova_literacy <- lrtest(full_model, no_literacy_model)

# Compare Full Model vs. Model without Region
anova_region <- lrtest(full_model, no_region_model)

# Display Results
print(anova_literacy)
print(anova_region)
print(anova_marriage)

```



# Conclusion
Our final model focus that both literacy and age at marriage are significant across regional contexts, with age at marriage having the most substantial impact on family size.

For an illiterate person married between the ages of 0-15, the family size difference between Lisbon and Porto is -0.18. Which indicate the ratio of family sizes between Lisbon and Porto is 0.835, holding other factors constant.

Considering the two largest cities in Portugal, Porto and Lisbon: for a literate individual in Lisbon, married between the ages of 22-25, the family size ratio is approximately 0.741 times that of an illiterate individual in Porto, also married between the ages of 22-25. This is calculated by multiplying the effects of literacy (0.888), region (0.835), and age at marriage (0.970). Thus, the combined ratio is approximately 0.741 (i.e., 0.888 × 0.835 × 0.970 = 0.741).

Our finding allign with previous research, education and marrage age will highly impact family size, which arise a issue with [write something related wit our citation].

# Appendix {-}

\newpage

# References {-}
